---
title: "Reproducable Research Project 2: Analysis of US Storm Data By Event Type and Impact"
author: "jwismer"
date: "Sunday, November 23, 2014"
output: html_document
---

## Synopsis
There is anecdotal evidence that the frequency and severity of storms have been increasing over time,
possibly due to global warming impact. 

Lessening the impact by careful planning may lower the severe nature of the outcomes.  

This analysis seeks to identify the most impactful types of weather related events based on
storm data information collected by the National Weather Service (NWS).

This study explores the U.S. National Oceanic Atmospheric Administration's (NOAA) storm
database.  This database tracks storm severty factors for major storms across the United States.

Data coverage of weather events in the database starts in 1950 and ends in November 2011, with coverage
of the more recents years being considered more complete.

This study looks at the following variables associated with each weather observation:

1. EVTYPE:      Catigorical variable identifying the type of weather related event
2. FATALITIES:  numerical variable providing the number of deaths due to the specific event
3. INJURIES:    numerical variable providing the number of injuries due to the specific event
4. PROPDMG:     numerical variable providing an estimate of the amount of property damage due to the specific event
5. CROPDMG:     numerical variable providing an estimate of the amount of crop damage due to the specific event

## Data Processing

### Load the database
Retrieve and load the NOAA database, from the URL:
https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

```{r, echo=TRUE}
# get the data file
url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"

# Load and preprocess the data
# assume zip file has been downloaded already
filename <- "./repdata-data-StormData.csv.bz2"

#download unzip the file if found
if (!file.exists(filename)) {
    download.file(url, filename)
}

#conditionally load the dataset
if (!exists("stormData")) {
    stormData <- read.csv(bzfile(filename))
}

#subset to just the factors needed for the analysis
eventImpact <- stormData[ , c("EVTYPE", "FATALITIES", "INJURIES", 
                               "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP")]
```

### Aggregate the Data by Weather Event Type
We will group the data by event type to sum over specific variables considered for the study.

```{r, echo=TRUE}
#aggregate factors associated the harmful impact, by event type
fatality_agg <- aggregate(FATALITIES ~ EVTYPE, data=eventImpact, FUN="sum")
injury_agg   <- aggregate(INJURIES   ~ EVTYPE, data=eventImpact, FUN="sum")
#aggregate factors associated the highest economic impact, by event type
propdmg_agg <- aggregate(PROPDMG ~ EVTYPE, data=eventImpact, FUN="sum")
cropdmg_agg   <- aggregate(CROPDMG   ~ EVTYPE, data=eventImpact, FUN="sum")
```

## Results

### Question 1
**Across the US, which type of events (as indicated by the EVTYPE variable) are
most harmfull with respect to population?**

Top 6 fatal event types:  
```{r, echo=TRUE}
fatality_agg[head(order(fatality_agg$FATALITIES, decreasing = TRUE)), ]
```

Top 6 injurous event types:
```{r, echo=TRUE}
injury_agg[head(order(injury_agg$INJURIES, decreasing = TRUE)), ]
```
Plot of Most Harmful Storm Types
```{r, echo=TRUE, fig.height=10, fig.width=10}
par(mfrow = c(2,1))
barplot(fatality_agg[head(order(fatality_agg$FATALITIES, decreasing = TRUE)), "FATALITIES"], 
        names.arg=fatality_agg[head(order(fatality_agg$FATALITIES, decreasing = TRUE)), "EVTYPE"],
        main="Total Fatalities by Weather Event Type: Top 6 Most Fatal", ylab="Total Fatalities")
barplot(injury_agg[head(order(injury_agg$INJURIES, decreasing = TRUE)), "INJURIES"], 
        names.arg=injury_agg[head(order(injury_agg$INJURIES, decreasing = TRUE)), "EVTYPE"],
        main="Total Injuries by Weather Event Type: Top 6 Most Injurous", ylab="Total Injuries")
```

As we can see from the above plot, the **Tornado** event type is by far the most harmful, 
within the US.

### Question 2
**Across the US, which types of events have the greatest economic impact?**

Top 6 property damaging event types
```{r, echo=TRUE}
propdmg_agg[head(order(propdmg_agg$PROPDMG, decreasing = TRUE)), ]
```

Top 6 crop damaging event types
```{r, echo=TRUE}
cropdmg_agg[head(order(cropdmg_agg$CROPDMG, decreasing = TRUE)), ]
```

Plot of Most Impactful Storm Types
```{r, echo=TRUE, fig.height=10, fig.width=10}
par(mfrow = c(2,1))
barplot(propdmg_agg[head(order(propdmg_agg$PROPDMG, decreasing = TRUE)), "PROPDMG"], 
        names.arg=propdmg_agg[head(order(propdmg_agg$PROPDMG, decreasing = TRUE)), "EVTYPE"],
        main="Total Property Damage by Weather Event Type: Top 6 Most Damaging", ylab="Total Damage")
barplot(cropdmg_agg[head(order(cropdmg_agg$CROPDMG, decreasing = TRUE)), "CROPDMG"], 
        names.arg=cropdmg_agg[head(order(cropdmg_agg$CROPDMG, decreasing = TRUE)), "EVTYPE"],
        main="Total Crop Damage by Weather Event Type: Top 6 Most Damaging", ylab="Total Damage")

```

The answer to this question is not clear cut.  If one is concerned about an urban area, then planning
for a **Tornadic** event would be more beneficial than planning for a **Hail** event.  
However, in a rural area where farming is occuring, concern for a Hail event may be more prudant.